### 2015-04-15 concl.tex

This paper presents \name, a shallow Haskell extension for declarative
programming over \ecds. The key idea underlying \name's design is the automatic
classification of fine-grained consistency contracts on operations and
distributed transactions with respect to the consistency and isolation levels
offered by the store. Our contract language is carefully crafted from a
decidable subset of first-order logic, enabling the use of automated
verification tools to discharge the proof obligations associated with contract
classification. We realize an instantiation of \name on top of off-the-shelf
distributed store, Cassandra, and illustrate the benefit of fine-grained
contract classification by implementing and evaluating several scalable
applications.

>>>>

This paper presents \name, a shallow Haskell extension for declarative
programming over \ecds. The key idea underlying \name's design is the automatic
classification of fine-grained consistency contracts on operations and
distributed transactions with respect various consistency and isolation levels
offered by the store. Our contract language is carefully crafted from a
decidable subset of first-order logic, enabling the use of automated
verification tools to discharge the proof obligations associated with contract
classification. We realize an instantiation of \name on top of an off-the-shelf
distributed store, Cassandra, and illustrate the benefit of fine-grained
contract classification by implementing and evaluating several scalable
applications.

### 2015-04-15 contract-language.tex

The theorem states that if a data store correctly enforces $\scc$, $\ccc$, and
$\ecc$ contracts in all well-formed executions, then the same store, extended
with the classification scheme shown in Figure~\ref{sem:classify}, can enforce
all well-formed \name contracts. The proof of the theorem is given in the
accompanying technical report~\cite{techrep} \footnote{The technical report also
provides the concrete reduction rules for enforcing the consistency classes.}.

>>>>

\noindent The theorem states that if a data store correctly enforces $\scc$,
$\ccc$, and $\ecc$ contracts in all well-formed executions, then the same
store, extended with the classification scheme shown in
Figure~\ref{sem:classify}, can enforce all well-formed \name contracts. The
proof of the theorem is given below:

\begin{proof}
  Hypothesis:
  \begin{smathpar}
  \begin{array}{cr}
    \auxred{} {\E,\langle op,\tau \rangle;\sigma \pll \Sigma} {\eff}
    {\E', \sigma \pll \Sigma} & H\npp\\
    \E' \models \cv_\tau[\eff/\cureff] & H\npp\\
  \end{array}
  \end{smathpar}
  Since $\tau$ is the contract class of $\cv$, by inversion, we have
  $\cv \le \cv_\tau$. By the definition of $\le$ relation:
  \begin{smathpar}
  \begin{array}{cr}
    \Delta \vdash \forall \cureff. \cv_\tau \Rightarrow \cv & H\npp\\
  \end{array}
  \end{smathpar}
   Since $\eff$ denotes new effect, it is a fresh variable that does
   not occur free in $\Delta$. From $H3$, after instantiating bound
   $\cureff$ with $\eff$, we have:
  \begin{smathpar}
  \begin{array}{cr}
    \Delta \vdash \cv_\tau[\eff/\cureff] \Rightarrow \cv[\eff/\cureff]
      & H\npp\\
  \end{array}
  \end{smathpar}
  Due to the soundness of natural deduction for first-order logic,
  $H4$ implies that for all models $\mathcal{M}$ such that
  $\mathcal{M} \models \Delta$, if $\mathcal{M} \models
  \cv_\tau[\eff/\cureff]$ then $\mathcal{M} \models
  \cv[\eff/\cureff]$. Since $\E'$ is well-formed, we have:
  \begin{smathpar}
  \begin{array}{cr}
    \E' \models \Delta & H\npp\\
  \end{array}
  \end{smathpar}
  Proof follows from $H1$, $H4$, and $H3$.
  \hfill \qed
\end{proof}

It is important to note that Theorem~\ref{thm:classification-sound}
does not ascribe any semantics to the reduction relation
($\xrightarrow{}$). As such, it makes no assumptions about how the
store actually implements ${\sf ec}$, ${\sf cc}$ and ${\sf sc}$
guarantees. The specific implementation strategy is determined by the
operational semantics of the store, which \emph{defines} the reduction
relation for that particular store. The following section describes
operational semantics of the store used by the \name implementation.

### 2015-04-15 implementation.tex
Edits to Lattice figure. Edits to caption for atomicity implementation.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{Figures/AtomicityImpl}
\end{center}
\caption{Implementing atomicity semantics. Dotted circle represents effects not
yet inserted into the backing store.}
\label{fig:atomicity_impl}
\end{figure}

\name implements atomic visibility by exploiting shim layer causality guarantees
-- an effect is included only if all the effects if depends on are also
included. Consider the example given in Figure~\ref{fig:atomicity_impl}. For
every transaction in \name, we instantiate a special transaction marker effect
$m$ that is importantly not inserted into the backing store. Marker $m$ is
included as a dependence to every effect generated in the transaction. In the
figure, the top graph on the left shows the state of the store in the middle of
a transaction. Each circle represents an effect. The dotted circle indicates
that the effect has been instantiated, but has not yet been inserted into the
store. Since the causally preceding effect $m$ has not yet been written to the
store, no operation will witness $e1$ and $e2$ while the transaction in
progress. After the transaction has finished execution, we insert $m$ into the
backing store, marking all the effects from the transactions as a dependence
for $m$, as show in the bottom graph. Now, any replica which includes one of
the effects from the transaction must include $m$, and transitively must
include every effect from the transaction. This ensures atomicity and satisfies
the RC requirement.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{Figures/SumDisk}
\end{center}
\caption{Summarization in the backing store. Dotted circle represents effects
not yet inserted into the backing store.}
\label{fig:impl_sum_disk}
\end{figure}

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=0.8\columnwidth]{Figures/AtomicityImpl}
\end{center}
\caption{Implementing atomicity semantics.}
\label{fig:atomicity_impl}
\end{figure}

\name implements atomic visibility by exploiting shim layer causality
guarantees -- an effect is included only if all the effects if depends on are
also included. Consider the example given in Figure~\ref{fig:atomicity_impl}.
In the figure, the graphs represent the state of the store, where the circles
represent effects and the edges represent the dependence between effects. The
dotted circle represents effects that are not yet inserted into the store.
The graph on the left shows that state of the store after executing \cf{oper2}.
For every transaction in \name, we instantiate a special transaction marker
effect $m$ that is importantly not inserted into the backing store. Marker $m$
is included as a dependence to every effect generated in the transaction. Since
the causally preceding effect $m$ has not yet been written to the store, no
operation will witness $e1$ and $e2$ while the transaction in progress. After
the transaction has finished execution, we insert $m$ into the backing store,
marking all the effects from the transactions as a dependence for $m$, as shown
in the graph on the right. Now, any replica which includes one of the effects
from the transaction must include $m$, and transitively must include every
effect from the transaction. This ensures atomicity and satisfies the RC
requirement.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{Figures/SumDisk}
\end{center}
\caption{Summarization in the backing store.}
\label{fig:impl_sum_disk}
\end{figure}

### 2015-04-15 related.tex

Our contract language and system model is inspired by the axiomatic
description of RDT semantics proposed in~\cite{Burckhardt2014}. While
~\cite{Burckhardt2014} uses axioms for formal verification of
an RDT implementation, we utilize them as a means for the user to express
desired consistency guarantees in the application. The operational semantics
of \name (described in the accompanying tech report~\cite{techrep}) describes
an operational manifestation of our system model. \cite{Burckhardt2015} also
presents an operational model of a replicated data store that is based on
the abstract system model presented in ~\cite{Burckhardt2014}. Their claims
about the expressivity and practicality of the system model vindicate
ours. However, both approaches differ in the way they aim to empower the application
programmer. \cite{Burckhardt2015} exposes the operational model to the
programmer, along with primitives, such as \cf{push}, \cf{pull}
and \cf{fetch}, that make it easy to \emph{implement} required
consistency and isolation guarantees. In contrast, \name completely
abstracts its operational model, and instead relies on contract
classification to automatically \emph{choose} appropriate consistency and isolation
levels. Similar to~\cite{Burckhardt2014} and~\cite{Burckhardt2015}, our
system model does not incorporate real (i.e., wall-clock) time. Hence, our
contract language cannot describe store semantics such as recency or
bounded-staleness guarantees offered by certain stores~\cite{Pileus}.

>>>>

Our contract language and system model is inspired by the axiomatic description
of RDT semantics proposed in~\cite{Burckhardt2014}. While ~\cite{Burckhardt2014}
uses axioms for formal verification of an RDT implementation, we utilize them as
a means for the user to express desired consistency guarantees in the
application. The operational semantics of \name describes an operational
manifestation of our system model.~\cite{Burckhardt2015} also presents an
operational model of a replicated data store that is based on the abstract
system model presented in ~\cite{Burckhardt2014}. Their claims about the
expressivity and practicality of the system model vindicate ours. However, both
approaches differ in the way they aim to empower the application programmer.
\cite{Burckhardt2015} exposes the operational model to the programmer, along
with primitives, such as \cf{push}, \cf{pull} and \cf{fetch}, that make it easy
to \emph{implement} required consistency and isolation guarantees. In contrast,
\name completely abstracts its operational model, and instead relies on contract
classification to automatically \emph{choose} appropriate consistency and
isolation levels.  Similar to~\cite{Burckhardt2014} and~\cite{Burckhardt2015},
our system model does not incorporate real (i.e., wall-clock) time. Hence, our
contract language cannot describe store semantics such as recency or
bounded-staleness guarantees offered by certain stores~\cite{Pileus}.

### 2015-04-15 transactions.tex

While contracts on individual operations offer the programmer object-level
declarative reasoning, real-world scenarios often involve operations that span
multiple objects. In order to address this problem, several recent
systems~\cite{Walter,Burckhardt2012,BailisHAT} have proposed eventually
consistent transactions in order to compose operations on multiple objects.
However, given that classical transaction models such as serializability and
snapshot isolation require inter-replica coordination, these systems espouse
\emph{coordination-free transactions} that remain available under network
partitions, but only provide weaker isolation guarantees. Coordination-free
transactions have intricate consistency semantics and widely varying runtime
overheads. As with operation-level consistency, the onus is on the programmer
to pick the correct transaction kind. This choice is further complicated by
consistency semantics of individual operations.

@@

\name automates the choice of assigning the correct and most efficient
transaction isolation level. Similar to contracts on individual operations, the
programmer associates contracts with transactions, declaratively expressing its
consistency specification. We extend the contract language with a new term
under quantifier-free propositions - ${\small \txnZ}~S_1~S_2$, where $S_1$ and
$S_2$ are sets of effects, and introduce a new primitive equivalence relation
$\small \sametxnZ$ that holds for effects from the same transaction. $\small
\txn{a,b}{c,d}$ is just syntactic sugar for $\small \sametxn{a}{b} ~\wedge~
\sametxn{c}{d} ~\wedge~ \neg\sametxn{a}{c}$, where $a$ and $b$ considered to
belong to the \emph{current} transaction.

@@

In order to illustrate the utility of declarative reasoning for transactions,
consider an extension of our running example to use two accounts (objects) --
current ($c$) and savings ($s$). Each account provides operations
\rcf{withdraw}, \rcf{deposit} and \rcf{getBalance}, with the same contracts as
defined previously. We consider two transactions -- \cf{save(amt)}, which
transfers \cf{amt} from current to savings, and \cf{totalBalance}, which
returns the sum of the balances of individual accounts. Our goal is to ensure
that \cf{totalBalance} returns the result obtained from a consistent snapshot
of the object states. The \name code for the transactions is given below:

@@

While making both transactions serializable would ensure correctness,
distributed stores rarely offer serializable transactions since it is
unavailable and hinders scalability~\cite{BailisHAT}. As we will see, these
transactions can be satisfied with much weaker isolation guarantees. Despite
the atomicity offered by the transaction, anomalies are still possible. For
example, the two \rcf{getBalance} operations in \cf{totalBalance} transactions
might be served by different replicas with distinct set of committed \cf{save}
transactions. If the first(second) \rcf{getBalance} operation witness a
\cf{save} transaction that is not witnessed by the second(first)
\rcf{getBalance} operation, then the balance returned will be less(greater) than
the actual balance. It is not immediately apparent which weakest isolation
guarantee will be sufficient to prevent the anomaly.

Instead, \name requires the programmer to simply state the consistency
requirement as a contract. Since we would like both the \rcf{getBalance}
operations to witness the same set of \cf{save} transactions, we define the
constraint on \cf{totalBalance} transaction $\psi_{tb}$ as:

@@

The \cf{save} transaction only needs to ensure that the two writes it performs
are made visible atomically. Since this is ensured by combining them in a
transaction, \cf{save} does not require any additional constraints, and
$\psi_1$ is simply $\small \true$.

@@

In order to illustrate the utility of transaction contract classification, we
identify three well-understood coordination-free transaction semantics -- Read
Committed (RC)~\cite{Berenson95}, Monotonic Atomic View (MAV)~\cite{BailisHAT}
and Repeatable Read (RR)~\cite{Berenson95}, and illustrate the classification
strategy. Our technique can indeed be applied to a different isolation level
lattice.

A transaction with ANSI RC semantics only witnesses committed operations. Let
us assume that a replica will buffer transactional updates until all the
updates from the same transaction are available at that replica. Once all the
updates from a transaction are available, the buffered updates are made visible
to subsequent client requests. This ensure atomicity of transactions.
Importantly, RC does not entail any other guarantees. As a result, a store
implementing RC does not require inter-replica coordination. We can express RC
as follows:

@@

\noindent Notice that the above definition is the same as the atomicity
guarantee of transaction described in \S~\ref{sec:syn_sem_ext}. The \cf{save}
is an example for RC transaction.

@@

ANSI RR semantics requires that the transaction witness a snapshot of the data
store state. Importantly, this snapshot can be obtained from any replica, and
hence RR is coordination-free. An example for such a transaction is the
\cf{totalBalance} transaction. The semantics of RR is captured by the following
contract:

>>>>

While contracts on individual operations offer the programmer object-level
declarative reasoning, real-world scenarios often involve operations that span
multiple objects. In order to address this problem, several recent
systems~\cite{Walter,Burckhardt2012,BailisHAT} have proposed eventually
consistent transactions in order to compose operations on multiple objects.
However, given that classical transaction models such as serializability and
snapshot isolation require inter-replica coordination, these systems espouse
\emph{coordination-free transactions} that remain available under network
partitions, but only provide weaker isolation guarantees. Coordination-free
transactions have intricate consistency semantics and widely varying runtime
overheads. As with operation-level consistency, the onus is on the programmer
to pick the correct transaction kind. This choice is further complicated by
the consistency semantics of individual operations.

@@

\name automates the choice of assigning the correct and most efficient
transaction isolation level. Similar to contracts on individual operations, the
programmer associates contracts with transactions, declaratively expressing 
consistency specifications. We extend the contract language with a new term
under quantifier-free propositions - ${\small \txnZ}~S_1~S_2$, where $S_1$ and
$S_2$ are sets of effects, and introduce a new primitive equivalence relation
$\small \sametxnZ$ that holds for effects from the same transaction. $\small
\txn{a,b}{c,d}$ is just syntactic sugar for $\small \sametxn{a}{b} ~\wedge~
\sametxn{c}{d} ~\wedge~ \neg\sametxn{a}{c}$, where $a$ and $b$ considered to
belong to the \emph{current} transaction.

@@

In order to illustrate the utility of declarative reasoning for transactions,
consider an extension of our running example to use two accounts (objects) --
current ($c$) and savings ($s$). Each account provides operations
\rcf{withdraw}, \rcf{deposit} and \rcf{getBalance}, with the same contracts as
defined previously. We consider two transactions -- \cf{save(amt)}, which
transfers \cf{amt} from current to savings, and \cf{totalBalance}, which
returns the sum of the balances of individual accounts. Our goal is to ensure
that \cf{totalBalance} returns the result obtained from a consistent snapshot
of the object states. The \name code for these transactions is given below:

@@

While making both transactions serializable would ensure correctness,
distributed stores rarely offer serializable transactions given their
availability requirements and implications for
scalability~\cite{BailisHAT}. Fortunately, these transactions can be
satisfied with much weaker isolation guarantees. However, despite the atomicity
offered by the transaction, anomalies are still possible. For example, the
two \rcf{getBalance} operations in a \cf{totalBalance} transaction might be
served by different replicas with a distinct set of committed \cf{save}
transactions. If the first(second) \rcf{getBalance} operation witnesses a
\cf{save} transaction that is not witnessed by the second(first)
\rcf{getBalance} operation, then the balance returned will be less(greater)
than the actual balance. It is not immediately apparent how to choose the weakest
isolation guarantee that would be sufficient to prevent the anomaly.

Instead, \name requires the programmer to simply state the consistency
requirement as a contract. Since we would like both the \rcf{getBalance}
operations to witness the same set of \cf{save} transactions, we define the
constraint on the \cf{totalBalance} transaction $\psi_{tb}$ as:

@@

The \cf{save} transaction only needs to ensure that the two writes it performs
are made visible atomically. Since this is ensured by combining them in a
transaction, \cf{save} does not require any additional constraints, and
$\psi_{sv}$ is simply $\small \true$.

@@

In order to illustrate the utility of transaction contract classification, we
identify three well-understood coordination-free transaction semantics -- Read
Committed (RC)~\cite{Berenson95}, Monotonic Atomic View (MAV)~\cite{BailisHAT}
and Repeatable Read (RR)~\cite{Berenson95}, and illustrate the classification
strategy. Our technique can indeed be applied to a different isolation-level
lattice.

A transaction with ANSI RC semantics only witnesses committed operations. Let
us assume that a replica will buffer transactional updates until all the
updates from the same transaction are available at that replica. Once all the
updates from a transaction are available, the buffered updates are made visible
to subsequent client requests. This ensures atomicity of transactions.
Importantly, RC does not entail any other guarantees. As a result, a store
implementing RC does not require inter-replica coordination. We can express RC
as follows:

@@

\noindent Notice that the above definition is the same as the atomicity
guarantee of transaction described in \S~\ref{sec:syn_sem_ext}. The \cf{save}
operation is an example of an RC transaction.

@@

ANSI RR semantics requires that the transaction witness a snapshot of the data
store state. Importantly, this snapshot can be obtained from any replica, and
hence RR is also coordination-free. An example for such a transaction is the
\cf{totalBalance} transaction. The semantics of RR is captured by the following
contract:
